---
title: "Reverse Engineering Project"
author: "Alex Taylor, Jordan Fine, Devin Etta"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [STORY NAME](STORY URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}

Donations <- read_csv("sports-political-donations.csv") |>
  clean_names() |>
  mutate(
    amount = as.numeric(gsub("[$,]", "", amount))  # removes $ and commas
  )
```


## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1 - Jordan

* **Sentence text**: “American professional sports owners have contributed nearly $47 million in federal elections since 2015… $34.2 million (72.9 percent) [went] to Republican campaigns or super PACs… compared to $10.1 million (21.5 percent) to Democrats.”

* **Analysis summary**: I think this quote includes an interesting statistic that provides the main point of the article. There is a big difference in how much money went to republicans vs democrats, this could mean something important or make for an important story. 


```{r}
# Split dataset by party
republican_donors <- sports_donations |> filter(party == "Republican")
democrat_donors   <- sports_donations |> filter(party == "Democrat")

# Calculate totals, averages, and counts for each party
party_summary <- sports_donations |>
  group_by(party) |>
  summarize(
    total_donated = sum(amount, na.rm = TRUE),
    average_donation = mean(amount, na.rm = TRUE),
    number_of_donors = n()
  )

# Calculate percentages of total donations
total_all <- sum(party_summary$total_donated)
party_summary <- party_summary |>
  mutate(percent_of_total = (total_donated / total_all) * 100)

# Arrange from highest to lowest total donated
party_summary <- party_summary |> arrange(desc(total_donated))

party_summary

 

```

### Sentence 2 - Alex

* **Sentence text**: Overall, 51.7 percent of all contributions by WNBA owners went to Democrats, with 42.3 percent going to Republicans.

* **Analysis summary**: I was not able to confirm these results. In my analysis, 76.6% of WNBA contributions went to Democrats, while 19.7% went to Republicans. At first, I thought this was because an owner owned a team in a league other than the WNBA, but after this ended up not being the case. 

```{r}
# Put code to reverse engineer sentence here

# Filter to only focus on WNBA players, then see how many contributions were made to each party
WNBA_Contribs <- Donations |>
  filter(league == "WNBA") |>
  group_by(party) |>
  summarise(numb_contrib = n())

# Find the percentage of contributions each party got
WNBA_Contribs <- WNBA_Contribs |>
  mutate(
    pct_contrib = numb_contrib / sum(numb_contrib) * 100
  )

# Display results of code below this codeblock
WNBA_Contribs

```

### Sentence 3 - Devin (Did 5th sentance in document)

* **Sentence text**: “Forty owners account for more than $40 million — more than 85 percent of all contributions. And looking specifically at these big spenders, it’s clear where their allegiances lie: 77.7 percent of that $40 million went to Republican causes or candidates, with 18.8 percent going to Democrats.” 
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

#The data needs to be clean like this because the amount data set needs to be set to numeric
Clean_Donations <- Donations |>
  clean_names() |>
  mutate(amount = as.numeric(gsub("[$,]","",amount))) 

Clean_Donations



#top 40 donations code  
top_40donations <- Clean_Donations |>
  group_by(owner) |>
  summarize(total_amount = sum(amount))|>
  top_n(40,total_amount) |>
  summarize(total_from40 = sum(total_amount))

top_40donations
   #calculate the total contributions       
 total_donation_amount <- sum(Clean_Donations$amount)
   
 #percentage 
 percent_top40 <- (top_40donations$total_from40/ total_donation_amount) * 100

# Display results of code below this codeblock
  
 #the percent of the top 40 owners 
 percent_top40

```


### Sentence 4

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: “Forty owners account for more than $40 million — more than 85 percent of all contributions. And looking specifically at these big spenders, it’s clear where their allegiances lie: 77.7 percent of that $40 million went to Republican causes or candidates, with 18.8 percent going to Democrats.” 
* **Analysis summary**:This paragraph displays a statistic that slices the data and looks at the top donors. The github data from the article helps to list the owners and their sport teams that have a concentrated spending power. The analysis could be easy by just filtering the minimum spending amount to only display those donors. Either the data can be grouped by donations, or just the owners. 




```{r}
owner_totals <- sports_donations |> 
  group_by(owner) |> 
  summarize(total_donated = sum(amount, na.rm = TRUE)) |> 
  arrange(desc(total_donated))
top40_owners <- owner_totals |> 
  slice_max(order_by = total_donated, n = 40)
sports_donations |> 
  semi_join(top40_owners, by = "owner") |> 
  group_by(party) |> 
  summarize(
    total_donated = sum(amount, na.rm = TRUE),
    percent = total_donated / sum(total_donated) * 100
  )
```

-30-